\documentclass{article}
\usepackage{amsmath}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
% \pgfplotsset{compat=1.14}
\usepgfplotslibrary{groupplots}
\usetikzlibrary{spy}
\usetikzlibrary{external} % Probaby doesn't work on overleaf :(
\usepackage{listings}
\usepackage{siunitx}
\usepackage{cleveref}
\usepackage{booktabs}
\usepackage{color} %red, green, blue, yellow, cyan, magenta, black, white
\definecolor{mygreen}{RGB}{28,172,0} % color values Red, Green, Blue
\definecolor{mylilas}{RGB}{170,55,241}

\lstset{language=Matlab,%
    %basicstyle=\color{red},
    breaklines=true,%
    morekeywords={matlab2tikz},
    keywordstyle=\color{blue},%
    morekeywords=[2]{1}, keywordstyle=[2]{\color{black}},
    identifierstyle=\color{black},%
    stringstyle=\color{mylilas},
    commentstyle=\color{mygreen},%
    showstringspaces=false,%without this there will be a symbol in the places where there is a space
    numbers=left,%
    numberstyle={\tiny \color{black}},% size of the numbers
    numbersep=9pt, % this defines how far the numbers are from the text
    emph=[1]{for,end,break},emphstyle=[1]\color{red}, %some words to emphasise
    %emph=[2]{word1,word2}, emphstyle=[2]{style},    
}

\pgfplotsset{/pgfplots/table/tableformat/.style={
  col sep=semicolon,  % the separator in our .csv file
  every head row/.style={before row=\toprule,after row=\midrule},
  every last row/.style={after row=\bottomrule},
  every column/.append style={numeric type, dec sep align},
  },
}

\def\d{\mathrm{d}}
\tikzexternalize

\date{2018-09-14}
\author{Jing Ying Ko jyko@kth.se 
\and Henrik Grimler hgrimler@kth.se}
\title{Computer exercise 3 in SF2520 \\Partial differential equation of parabolic type \\CE3 37 }
\begin{document}

\maketitle
\section{Rescaling to dimensionless form}

The following partial differential equation is given:
\begin{equation} \label{eq1}
\rho c_p \frac{\partial T}{\partial t} = \kappa \frac{\partial^2 T}{\partial x^2}
\end{equation}
with boundary conditions
\begin{equation} \label{eq2}
   T(0,t)= 
\begin{cases}
    T_0,           & 0 \leq t \leq t_P, \\
    0,              & t > t_P,
\end{cases}
   \hspace{0.8cm}         \frac{\partial T}{\partial x}(L,t)=0,
\end{equation}
and initial conditions
\begin{equation} \label{eq3}
   T(x,0)= 
\begin{cases}
    T_0,           & x=0, \\
    0,              & 0 < x \leq L
\end{cases}
\end{equation}
where $\rho$ is the density [\si{\kilo\gram\per\meter\cubed}], $c_p$ is the
heat capacity [\si{\joule\per\kilo\gram\per\kelvin}] and
$\kappa$ is the thermal conductivity [\si{\joule\per\meter\per\second\per\kelvin}] of the rod.
\newline

\noindent
1. Introduce reference values ($T_0$,$L$,$t_P$) for each of the three variables ($T,x,t$) and define the corresponding dimensionless variables
\begin{equation}
    T = T_0 u, \hspace{0.8cm} x = L \xi, \hspace{0.8cm} t = t_P \tau.
\end{equation}

\noindent
2. Rewrite differential equations in those variables
\begin{equation}
\begin{split}
    \frac{\partial T}{\partial t} &= \frac{T_0}{t_P} \frac{\partial u}{\partial \tau} \\
    \frac{\partial}{\partial x} &= \frac{1}{L} \frac{\partial}{\partial \xi} \\
    \frac{\partial^2 T}{\partial x^2} &= \frac{T_0}{L^2} \frac{\partial^2 u}{\partial \xi^2}
\end{split}
\end{equation}
Rewrite \cref{eq1} into
\begin{equation}
\begin{split}
    \rho c_p \frac{\partial u}{\partial \tau} &= \kappa \frac{T_0}{L^2} \frac{\partial^2 u}{\partial \xi^2} \\
    \frac{\partial u}{\partial \tau} &= \frac{\kappa t_p}{\rho c_p L^2} \frac{\partial^2 u}{\partial \xi^2}
\end{split}
\end{equation}
and therefore can be transformed into the following dimensionless form
\begin{equation}
    \frac{\partial u}{\partial \tau} = a \frac{\partial^2 u}{\partial \xi^2}, \hspace{0.8cm} \tau > 0, \hspace{0.8cm} 0 < \xi < \frac{L}{L}=1
\end{equation}
with boundary conditions
\begin{equation}   \label{BC}
   u(0,\tau)= 
\begin{cases}
    \frac{T_0}{T_0} = 1,           & 0 \leq \tau \leq \frac{t_P}{t_P}, \\
    0,              & \tau > 1,
\end{cases}
   \hspace{0.8cm}         \frac{\partial u}{\partial \xi}(1,\tau) = 0,
\end{equation}
and initial conditions
\begin{equation} 
   u(\xi,0) = 
\begin{cases}
    \frac{T_0}{T_0} = 1,           & \xi = 0, \\
    0,              & 0 < \xi \leq \frac{L}{L}=1
\end{cases}
\end{equation}
with expression $a$ as
\begin{equation}
\begin{split}
    a &= \frac{\kappa t_P}{\rho c_p L^2} \\
    [a] &= \si{\joule\per\meter\per\second\per\kelvin\second\meter\cubed\per\kilo\gram\kilo\gram\kelvin\per\joule}
    = 1
\end{split}
\end{equation}
and assuming $a$ has a numerical value of 1.

\section{Discretization}

The semidiscretization approach is used. Using constant stepsize $h$ and space discretization into $n$ space:
\begin{equation}
    \xi_i = i \cdot h \hspace{0.8cm} \text{where}  \hspace{0.8cm} 0 \leq i \leq n+1 
\end{equation}
The discretized grid of the $\xi$-axis and how the gridpoints are numbered are
shown in \cref{discret}. $\xi_{0}=\alpha(\tau)$ is the boundary condition used at
the left-most point and is not part of the $u$-vector that is solved for. 
\tikzstyle{circ} = [circle, draw, fill=blue!20]
\tikzstyle{line} = [draw, -latex']
\begin{figure}[!h]
  \centering
  \begin{tikzpicture}[node distance = 1.5cm, auto]
    \node [circ              , label={below:$\xi_{1}$}] (xi1) {};
    \node [circ, left of=xi1 , label={below:$\xi_{0}$}] (xi0) {};
    \node [circ, right of=xi1, label={below:$\xi_{2}$} ] (xi2)   {};
    \node [circ, right of=xi2, label={below:$\ldots$} ] (xii)   {};
    \node [circ, right of=xii, label={below:$\xi_{i}$}] (xi3) {};
    \node [circ, right of=xi3  , label={below:$\ldots$}] (xi4) {};
    \node [circ, right of=xi4  , label={below:$\xi_{N}=1$}] (xin) {};
    \node [circ, right of=xin  , label={below:$\xi_{N+1}$}] (xin+1) {};
    \draw (xi1) -- (xi2) -- (xii) -- (xi3) -- (xi4) -- (xin);
    \draw[dashed] (xin) -- (xin+1);
    \draw[dashed] (xi1) -- (xi0);
  \end{tikzpicture}
  \caption{Schematic overview of discretized grids used in the method.}
  \label{discret}
\end{figure}

We approximate
\begin{equation}
    u_i(\tau) \approx u(\xi_i,\tau)
\end{equation}
and discretize $u_{\xi \xi}$ with approximation of order 2
\begin{equation}
    u_{\xi \xi}(\tau,\xi_i) \approx \frac{u_{i-1}(\tau) - 2u_i(\tau) + u_{i+1}(\tau)}{h^2} + 0(h^2) \hspace{0.8cm} i = 1, ..., n
\end{equation}
and so
\begin{equation}
    \frac{d u_i}{d\tau} = \frac{u_{i-1} - 2u_i + u_{i+1}}{h^2}  \hspace{0.8cm} i = 2, ..., n-1
\end{equation}
As for the boundary conditions at $i = 1$
\begin{equation}
    \frac{d u_1}{d\tau} = \frac{\alpha(\tau) - 2u_1 + u_{2}}{h^2}  \hspace{0.8cm} i = 1
\end{equation}
where 
\begin{equation} 
  \alpha(\tau)= 
\begin{cases}
    1,           & 0 \leq \tau \leq 1, \\
    0,              & \tau > 1,
\end{cases}
\end{equation}
and at $i = n$, a ghost point is used, and so
\begin{equation}
    \frac{d u_n}{d\tau} = \frac{u_{n-1} - 2u_n + u_{n+1}}{h^2}  \hspace{0.8cm} i = n
\end{equation}
Given the Newman boundary condition
\begin{equation}
    \frac{\partial u}{\partial \xi} (1,\tau) = 0
\end{equation}
which can be discretized in the form
\begin{equation}
\begin{split}
        \frac{\partial u}{\partial \xi} (1,\tau) &= \frac{u(1+h,\tau)-u(1-h,\tau)}{2h} + 0(h^2)\\
         \frac{\partial u}{\partial \xi} (1,\tau) &= \frac{u_{n+1}-u_{n-1}}{2h} + 0(h^2)
\end{split}
\end{equation}
and so
\begin{equation}
\begin{split}
    \frac{u_{n+1}-u_{n-1}}{2h} &= 0 \\
    u_{n+1} &= u_{n-1}
\end{split}
\end{equation}
resulting in
\begin{equation}
    \frac{d u_n}{d\tau} = \frac{2u_{n-1} - 2u_n}{h^2}  \hspace{0.8cm} i = n
\end{equation}

\noindent
Therefore, the equations that we need are
\begin{equation}
\begin{split}
    \frac{d u_1}{d\tau} &= \frac{\alpha(\tau) - 2u_1 + u_{2}}{h^2}  \hspace{0.8cm} i = 1 \\
    \frac{d u_i}{d\tau} &= \frac{u_{i-1} - 2u_i + u_{i+1}}{h^2}  \hspace{0.8cm} i = 2, ..., n-1    \\
    \frac{d u_n}{d\tau} &= \frac{2u_{n-1} - 2u_n}{h^2}  \hspace{0.8cm} i = n
\end{split}
\end{equation}
These can be written in the form
\begin{equation}
    \frac{d u}{d \tau} = Au + b(\tau), \hspace{0.8cm} u(0) = u_0
\end{equation}
where A is a $n$ x $n$ matrix, u is a $n$ x 1 vector, b($\tau$) is a $n$ x 1 vector and $u_0$ is a $n$ x 1 vector:
\begin{equation}
    A = \frac{1}{h^2} 
\begin{bmatrix}
    -2     &  1     &  0     &  0     & \dots  &  0     \\
    1      & -2     &  1     &  0     & \dots  &  0     \\
    0      &  1     & -2     &  1     & \dots  &  0     \\
    \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
    0      & \dots  &  1     & -2     &  1     &  0     \\
    0      & \dots  &  0     &  1     & -2     &  1     \\
    0      & \dots  &  0     &  0     &  2     & -2
\end{bmatrix}
\end{equation}
and
\begin{equation}
  b(\tau) = \frac{1}{h^2} 
  \begin{bmatrix}
    \alpha(\tau) \\
    0                        \\
    0                        \\
    \vdots                   \\
    0
  \end{bmatrix}
\end{equation}
and
\begin{equation}
  u_0 =
  \begin{bmatrix}
  % 1      \\ % u(0,\tau) isn't part of the u vector so shouldn't be included
  % in u_0?
    0      \\
    0      \\
    \vdots \\
    0
  \end{bmatrix}
\end{equation}


\section{Results}

The code to solve the described problem is in Appendix. For parabolic problems solved with explicit method requires the stability condition:
\begin{equation}
    \Delta t \leq 0.5 \cdot \Delta x^2
\end{equation}
Semidiscretizations of parabolic equations are stiff ODEs, as the timestep is severely restricted, i.e. since stepsize is already small, the timestep has to be much smaller. \cref{stable_solution_0} and \cref{stable_solution} show the stable solution while \cref{unstable_solution} show the unstable solution. A stepsize $h$ of 0.1 is chosen. In the stable solutions, we assign $\Delta t / h^2 = 0.3$ and $0.5$ respectively so the timesteps are $\Delta t =  0.003$ and $0.005$. We can see that both the ratios of $\Delta t / h^2$ of less than as well as equals 0.5 produce stable solutions. In the unstable solution, we assign $\Delta t / h^2 = 1.0$ so the timestep is $\Delta t =  0.01$.  \\ 

\noindent
\cref{2dplots} shows the temperature profiles along the rod when $\tau = 0.5, 1, 1.5, 2$. For $\tau$ up to 1, the temperature increases; whereas for $\tau > 1$, the temperature decreases. The behavior complies with the boundary conditions described in \cref{BC}.

\begin{figure}
  \centering
  \begin{tikzpicture}
    \begin{axis}
      [
        xlabel style = {sloped},
        ylabel style = {sloped},
        xlabel = {Dimensionless time},
        ylabel = {Dimensionless space},
        zlabel = {Dimensionless temperature},
      ]
      \addplot3 [surf, mesh/ordering=y varies] table {part3_stable_plot_0.csv};
    \end{axis}
  \end{tikzpicture}
  \caption{Stable plot with $\Delta t = 0.003$, $\Delta x = 0.1$ and the
    Courant number $0.3$.}
    \label{stable_solution_0}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \begin{axis}
      [
        xlabel style = {sloped},
        ylabel style = {sloped},
        xlabel = {Dimensionless time},
        ylabel = {Dimensionless space},
        zlabel = {Dimensionless temperature},
      ]
      \addplot3 [surf, mesh/ordering=y varies] table {part3_stable_plot.csv};
    \end{axis}
  \end{tikzpicture}
  \caption{Stable plot with $\Delta t = 0.005$, $\Delta x = 0.1$ and the
    Courant number $0.5$.}
    \label{stable_solution}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \begin{axis}
      [
        xlabel style = {sloped},
        ylabel style = {sloped},
        xlabel = {Dimensionless time},
        ylabel = {Dimensionless space},
        zlabel = {Dimensionless temperature},
      ]
      \addplot3 [surf, mesh/ordering=y varies] table {part3_unstable_plot.csv};
    \end{axis}
  \end{tikzpicture}
  \caption{Unstable plot with $\Delta t = 0.01$, $\Delta x = 0.1$ and the
    Courant number $1$.}
    \label{unstable_solution}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \begin{axis}
      [
        legend style = {at = {(0, 0.5)}, anchor = west},
        xlabel = {Dimensionless space},
        ylabel = {Dimensionless temperature},
      ]
      \addplot table [y index = 1] {part3_2D.csv};
      \addplot table [y index = 2] {part3_2D.csv};
      \addplot table [y index = 3] {part3_2D.csv};
      \addplot table [y index = 4] {part3_2D.csv};
      \legend{$\tau = 0.5$, $\tau = 1.0$, $\tau = 1.5$, $\tau = 2.0$}
    \end{axis}
  \end{tikzpicture}
  \caption{Solution at $\tau = 0.5$, $\tau = 1$, $\tau = 1.5$, $\tau = 2$}
    \label{2dplots}
\end{figure}

\section{Using MATLAB Solvers}

The ODE problem is then solved with two MATLAB solvers: the explicit method
\texttt{ode23} and the implicit method/stiff method \texttt{ode23s} which is
suitable for stiff problems. Three stepsizes $h$ are taken corresponding to
gridpoints $N = 10, 20, 40$ and $\tau$ of 0 to 2 is the time interval. The
following table is the comparison of number of time steps, CPU time and
maximal stepsize each method would take.

The number of time steps for \texttt{ode23s} is smaller than for
\texttt{ode23}. \texttt{ode23} requires more time steps for large numbers of
stepsizes to fullfill the stability criteria. On the other hand,
\texttt{ode23s} is unconditionally stable for every time step. Therefore, the
maximum time stepsize is bigger than the non-stiff method \texttt{ode23}.
\begin{table}[!h]
 \caption{Comparison of ode23, ode23s and the ode23s solution using a sparse
   Jacobian. The relative tolerance is set to $10^{-6}$ and the absolute
   tolerance to $10^{-6}$ for all solutions.}
 \label{tab:solution}
 \pgfplotstableread[col sep=semicolon, string type]{part4_solver_comparison.csv}{\partfour}
 \pgfplotstabletranspose[colnames from=N, input colnames to=N,string type]\partfourtransposed{\partfour}
 \centering\pgfplotstabletypeset[every nth row={3[-1]}{after row=\midrule},
   every head row/.style={
     before row=\toprule,
     after row=\midrule
   }, string type, skip first n=4,
   every last row/.style={
     after row=\bottomrule
     },]{\partfourtransposed}
\end{table}

\section{Optimized MATLAB Solvers}

\texttt{ode23s} is more efficient in solving the stiff problem as compared to
\texttt{ode23}. \texttt{ode23s} solved a linear systems of equations in order
to compute the next time step. In our problem, we have a system matrix of
equations which is tridiagonal, and so it should be faster to solve the linear
system. However, this is not taken advantage of when implementing
\texttt{ode23s}. Therefore, we can set an option to \texttt{ode23s} with
\texttt{odeset} to change the way the linear system is being solved. We can
either use '\texttt{Jacobian}' or '\texttt{Jpattern}'. '\texttt{Jacobian}'
can be used to specify a function that calculates the jacobian as a function
of time and the spatial variable, while '\texttt{Jpattern}' can be used to
give a sparse matrix containing the jacobian entries.

The number of time steps and the maximal stepsize are the same, only the CPU
time would change. \cref{tab:solution} shows that the CPU time is considerably
smaller when using ode23s with the '\texttt{Jpatten}' sparse matrix, compared
to when using ode23 out of the box.

\section{Conclusion}

The conclusions that can be drawn are:
\begin{itemize}
    \item Stiff method or implicit method (in this case \texttt{ode23s})
      should be used for parabolic problems as compared to non-stiff or
      explicit method. The semi-discretization using center difference method
      in space discretization of order 2 yields a linear system oof ODEs. The
      resulting matrix is a tridiagonal matrix with very different sizes of
      real eigenvalues. That is why, stiff method is better to solve this kind
      of parabolic problems when method of lines is used.
      
    \item The jacobian of the ODE-system is a sparse banded matrix. Most of
      the elements are zero (therefore, sparse). The non-zero entities are
      confined to a diagonal band and the bandwidth is relatively small
      (therefore, band/banded), in this case a tridiagonal matrix. That is,
      due to the center difference method, each point on the discretized space
      takes into account the neighbouring points (before and after). Being a
      sparse banded tridiagonal matrix, it is important that this structure is
      taken into consideration when solving for high efficiency. Using
      \texttt{Jacobian} and \texttt{Jpattern}, the solving time can be reduced
      as the jacobian is given and no computational effort has to be put to
      solve it numerically. 
\end{itemize}

\section{Appendix}

\subsection{main.m}
\lstinputlisting{main.m}
\subsection{build\_A\_du.m}
\lstinputlisting{build_A_du.m}
\subsection{print3\_data.m}
\lstinputlisting{print3_data.m}
\subsection{print\_data.m}
\lstinputlisting{print_data.m}
\subsection{custom\_fprintf.m}
\lstinputlisting{custom_fprintf.m}

\end{document}

